---
title: "Brazil Radar Data - Initial Exploration"
author: "PRU Analysis"
date: today
format:
  html:
    code-fold: true
    toc: true
---

## Data Overview

### Source Dataset

- **File**: `radar_2025_12_10.csv`
- **Size**: 1.3 GB, 8.8 million position reports
- **Coverage**: Single day (2025-12-10)
- **Format**: Semicolon-delimited CSV

### Data Fields

- **Identification**: `id`, `callsign`, `nr_ssr` (SSR code)
- **Route**: `addep`, `addes` (departure/arrival aerodromes)
- **Time**: `dh_inicio`, `dh_fim` (flight start/end), `dt_radar` (position timestamp)
- **Position**: `vl_latitude`, `vl_longitude`, `nr_flightlevel`
- **Dynamics**: `nr_speed` (ground speed)
- **Aircraft**: `ds_acctypes` (aircraft type list)

### Reference Data

- **Airport database**: 2,878 Brazilian aerodromes from OurAirports
- **ICAO prefixes**: SB, SD, SI, SJ, SN, SS, SW
- **Fields**: ICAO code, name, coordinates, elevation

## Setup

```{r}
#| label: setup
#| message: false

library(conflicted)
library(tidyverse)
library(arrow)
library(trrrj)
library(sf)
library(units)

conflict_prefer("filter", "dplyr")
conflicts_prefer(dplyr::lag)

# Path to radar_brazil utilities
radar_utils_path <- "../radar_brazil"

# Load local helper functions
source("R/sf_helpers.R")
```

## Load Data

```{r}
#| label: load-airports

brazil_airports <- file.path(radar_utils_path, "data-raw", "brazil_airports_arp.csv") |>
  read_csv(show_col_types = FALSE)
```

```{r}
#| label: define-schema

col_specs <- cols(
  id = col_double(),
  callsign = col_character(),
  addep = col_character(),
  addes = col_character(),
  dh_inicio = col_datetime(format = ""),
  dh_fim = col_datetime(format = ""),
  ds_acctypes = col_character(),
  vl_latitude = col_double(),
  vl_longitude = col_double(),
  nr_flightlevel = col_double(),
  nr_speed = col_double(),
  nr_ssr = col_double(),
  dt_radar = col_datetime(format = "")
)
```


## Extract Aerodrome Pair Trajectories

Target aerodrome pairs for initial HFE validation (bidirectional):

```{r}
#| label: define-pairs

# Aerodrome pairs to extract
target_pairs <- tribble(
  ~adep, ~ades,
  "SBCF", "SBKP",
  "SBKP", "SBCF",
  "SBCF", "SBSP",
  "SBSP", "SBCF",
  "SBBR", "SBCT",
  "SBCT", "SBBR",
  "SBBR", "SBSV",
  "SBSV", "SBBR",
  "SBGR", "SBSV",
  "SBSV", "SBGR",
  "SBGR", "SBRJ",
  "SBRJ", "SBGR"
)

target_pairs
```

```{r}
#| label: load-radar-sample

# Load full radar dataset (8.8M records)
df_raw <- read_delim(
  file = file.path(radar_utils_path, "data-raw", "radar_2025_12_10.csv"),
  delim = ";",
  trim_ws = TRUE,
  col_types = col_specs
)
```

```{r}
#| label: filter-trajectories

# Filter for target aerodrome pairs
df_filtered <- df_raw |>
  semi_join(target_pairs, by = c("addep" = "adep", "addes" = "ades"))

# Check coverage
df_filtered |>
  count(addep, addes) |>
  arrange(addep, addes)
```

## Trajectory Segmentation

Segment continuous radar position reports into discrete flight trajectories:

- **Temporal gap**: Split when time between positions > 30 minutes
- **Callsign change**: Split when callsign changes
- **Consistency filter**: Keep only flights with single consistent ADEP/ADES pair

```{r}
#| label: segment-trajectories

# Segment continuous radar tracks into discrete flights
df_trajectories <- df_filtered |>
  rename(
    pos_id = id,
    timestamp = dt_radar,
    adep = addep,
    ades = addes
  ) |>
  mutate(altitude = 100 * nr_flightlevel) |>
  select(
    pos_id, callsign, timestamp,
    longitude = vl_longitude,
    latitude = vl_latitude,
    altitude, adep, ades,
    everything()
  ) |>
  arrange(callsign, timestamp) |>
  mutate(
    delta = if_else(
      (!is.na(lag(callsign)) & callsign != lag(callsign)) |
        (!is.na(lag(timestamp)) & (timestamp - lag(timestamp)) > minutes(30)),
      1, 0
    ),
    fid = cumsum(delta)
  ) |>
  relocate(fid, .before = pos_id) |>
  select(-starts_with("dh_"), -starts_with("nr_"), -delta) |>
  group_by(fid) |>
  mutate(
    adep_n = n_distinct(adep, na.rm = TRUE),
    ades_n = n_distinct(ades, na.rm = TRUE)
  ) |>
  filter(adep_n == 1, ades_n == 1) |>
  ungroup() |>
  select(-ends_with("_n")) |>
  group_by(fid) |>
  fill(adep, .direction = "downup") |>
  fill(ades, .direction = "downup") |>
  ungroup() |>
  left_join(
    brazil_airports,
    by = c("adep" = "icao"),
    suffix = c("", "_adep")
  ) |>
  rename(name_adep = name, elevation_adep = elevation) |>
  left_join(
    brazil_airports,
    by = c("ades" = "icao"),
    suffix = c("", "_ades")
  ) |>
  rename(name_ades = name, elevation_ades = elevation)
```

## Summary Statistics

```{r}
#| label: summary-stats

# Flights per aerodrome pair
flight_summary <- df_trajectories |>
  group_by(adep, ades, name_adep, name_ades) |>
  summarise(
    n_flights = n_distinct(fid),
    n_positions = n(),
    .groups = "drop"
  ) |>
  arrange(adep, ades)

flight_summary
```

```{r}
#| label: trajectory-stats

# Overall extraction results
list(
  total_flights = n_distinct(df_trajectories$fid),
  total_positions = nrow(df_trajectories),
  aerodrome_pairs = nrow(flight_summary),
  date_coverage = range(df_trajectories$timestamp)
)
```

## Export Sample Dataset

Save extracted trajectories for ongoing HFE analysis work:

```{r}
#| label: export-data

# Create data directory
dir.create("data", showWarnings = FALSE)

# Save as parquet for efficient storage/loading
df_trajectories |>
  write_parquet("data/sample_trajectories_2025_12_10.parquet")

# Also save flight summary
flight_summary |>
  write_csv("data/flight_summary.csv")
```

**Output**:

- `data/sample_trajectories_2025_12_10.parquet`: All position reports for extracted flights
- `data/flight_summary.csv`: Summary table (flights and positions per pair)

## Flight Count Summary

```{r}
#| label: flight-count-summary

# Summary of detected movements per aerodrome pair
movement_summary <- df_trajectories |>
  group_by(adep, ades, name_adep, name_ades) |>
  summarise(
    flights = n_distinct(fid),
    positions = n(),
    .groups = "drop"
  ) |>
  arrange(desc(flights))

movement_summary |>
  select(
    `Origin` = adep,
    `Departure Airport` = name_adep,
    `Destination` = ades,
    `Arrival Airport` = name_ades,
    `Flights` = flights,
    `Positions` = positions
  ) |>
  knitr::kable(
    caption = "Detected movements for target aerodrome pairs (2025-12-10)",
    format.args = list(big.mark = ",")
  )
```

## Brazil HFE Reference Data

Load and subset the Brazilian HFE dataset for comparison:

```{r}
#| label: load-brazil-hfe

# Load full year Brazil HFE data
bra_hfe_full <- read_delim(
  file = file.path(radar_utils_path, "data-raw", "BRA-KEA-2025.csv"),
  delim = ";",
  trim_ws = TRUE,
  show_col_types = FALSE
)

# Filter for 2025-12-10 (our surveillance data date)
bra_hfe_dec10 <- bra_hfe_full |>
  mutate(date = as.Date(time)) |>
  filter(date == as.Date("2025-12-10"))

# Further filter for our target aerodrome pairs
bra_hfe_sample <- bra_hfe_dec10 |>
  semi_join(target_pairs, by = c("plan_dep" = "adep", "plan_arr" = "ades"))

# Summary
list(
  total_flights_year = nrow(bra_hfe_full),
  flights_dec10 = nrow(bra_hfe_dec10),
  flights_sample_pairs = nrow(bra_hfe_sample)
)
```

```{r}
#| label: brazil-hfe-sample-summary

# Show sample flights per aerodrome pair
bra_hfe_counts <- bra_hfe_sample |>
  count(plan_dep, plan_arr, name = "bra_hfe_flights")

bra_hfe_counts |>
  left_join(
    movement_summary |> select(adep, ades, flights),
    by = c("plan_dep" = "adep", "plan_arr" = "ades")
  ) |>
  rename(
    ADEP = plan_dep,
    ADES = plan_arr,
    `Brazil HFE` = bra_hfe_flights,
    `Our Extract` = flights
  ) |>
  knitr::kable(
    caption = "Flight counts comparison: Brazil HFE vs Our Radar Extract (2025-12-10)"
  )
```

```{r}
#| label: export-brazil-hfe-sample
#| eval: false

# Export Brazil HFE subset for ongoing comparison (disabled - file locked)
bra_hfe_sample |>
  write_csv("data/brazil_hfe_2025_12_10.csv")
```

**Output**: `data/brazil_hfe_2025_12_10.csv` - Brazil HFE reference for 2025-12-10 sample pairs (export disabled)

## Horizontal Flight Efficiency (HFE) Algorithm

### Methodology Overview

HFE measures the efficiency of the en-route portion of a flight by comparing the actual flown trajectory to the great circle distance between defined milestones.

**Key concepts** (see [ANS Performance Framework methodology](https://ansperformance.eu/methodology/horizontal-flight-efficiency-pi/)):

- **Milestones**: Fixed distance points from departure/to arrival (40 NM and 100 NM)
- **En-route segment**: Portion of flight between departure and arrival milestones
- **HFE indicator**: Ratio of great circle distance to actual flown distance (× 100%)

**Standard milestone pairs**:

- **40 NM**: Start at 40 NM from ADEP, end at 40 NM to ADES
- **100 NM**: Start at 100 NM from ADEP, end at 100 NM to ADES

### Milestone Extraction

Milestones are defined by **great circle distance from aerodrome reference points**:

- Distance FROM ADEP: GC distance from departure airport position
- Distance TO ADES: GC distance to arrival airport position
- DIST_FLOWN_NM: Cumulative distance along actual trajectory (from first point, in NM)

```{r}
#| label: calculate-milestones

# Function to calculate great circle distance (Haversine) — returns NM
great_circle_distance <- function(lat1, lon1, lat2, lon2) {
  R_nm <- 3440.065  # Earth radius in NM

  lat1_rad <- lat1 * pi / 180
  lat2_rad <- lat2 * pi / 180
  dLat <- (lat2 - lat1) * pi / 180
  dLon <- (lon2 - lon1) * pi / 180

  a <- sin(dLat/2)^2 + cos(lat1_rad) * cos(lat2_rad) * sin(dLon/2)^2
  c <- 2 * atan2(sqrt(a), sqrt(1-a))

  d <- R_nm * c
  return(d)
}

# Add GC distances from/to aerodromes and find milestones
df_with_gc <- df_trajectories |>
  group_by(fid) |>
  arrange(timestamp) |>
  cumulative_distance() |>
  mutate(
    # cumulative_distance from trrrj is in km — convert to NM
    cum_dist = cumulative_distance / 1.852,
    # GC distance from ADEP ARP (NM)
    gc_from_adep = great_circle_distance(
      latitude_adep, longitude_adep,
      latitude, longitude
    ),
    # GC distance to ADES ARP (NM)
    gc_to_ades = great_circle_distance(
      latitude, longitude,
      latitude_ades, longitude_ades
    )
  ) |>
  ungroup()

# Extract milestones for each flight and milestone type
extract_all_milestones <- function(.df) {
  milestone_list <- list()

  # FIRST_HIT: First trajectory point (proxy for takeoff)
  first_hit <- .df |>
    slice_head(n = 1) |>
    mutate(mst = "FIRST_HIT")
  milestone_list[[length(milestone_list) + 1]] <- first_hit

  # v1 milestones: 40NM from ADEP, 40NM to ADES (symmetric)
  for (target_nm in c(40, 100)) {

    # FROM ADEP: First point where distance FROM ADEP >= target (NM)
    ms_from <- .df |>
      filter(gc_from_adep >= target_nm) |>
      slice_head(n = 1) |>
      mutate(mst = paste0(target_nm, "NM_ADEP"))

    if (nrow(ms_from) > 0) {
      milestone_list[[length(milestone_list) + 1]] <- ms_from
    }

    # TO ADES: Last crossing into the target circle — last point where
    # gc_to_ades >= target AND the next point is already inside (< target).
    # This handles indirect routes / holdings where gc_to_ades oscillates.
    ms_to <- .df |>
      arrange(timestamp) |>
      mutate(next_gc_to_ades = lead(gc_to_ades)) |>
      filter(gc_to_ades >= target_nm,
             is.na(next_gc_to_ades) | next_gc_to_ades < target_nm) |>
      slice_tail(n = 1) |>
      mutate(mst = paste0(target_nm, "NM_ADES"))

    if (nrow(ms_to) > 0) {
      milestone_list[[length(milestone_list) + 1]] <- ms_to
    }
  }

  # Note: V2 reuses 40NM_ADEP + 100NM_ADES (no extra milestone needed)

  # LAST_HIT: Last trajectory point (proxy for landing)
  last_hit <- .df |>
    slice_tail(n = 1) |>
    mutate(mst = "LAST_HIT")
  milestone_list[[length(milestone_list) + 1]] <- last_hit

  bind_rows(milestone_list)
}

# Extract all milestones
df_milestones <- df_with_gc |>
  group_by(fid) |>
  group_modify(~ extract_all_milestones(.x)) |>
  ungroup() |>
  select(
    FLTID = fid,
    ADEP = adep,
    ADES = ades,
    TIME = timestamp,
    LAT = latitude,
    LON = longitude,
    ALT = altitude,
    DIST_FLOWN_NM = cum_dist,
    MST = mst,
    GC_FROM_ADEP_NM = gc_from_adep,
    GC_TO_ADES_NM = gc_to_ades
  )

# Preview milestone table
df_milestones |>
  arrange(FLTID, MST) |>
  head(20) |>
  knitr::kable(
    digits = 2,
    caption = "Milestone table (first 20 rows)"
  )
```

```{r}
#| label: check-milestones

# Verify which MST labels exist and how many flights have each
df_milestones |>
  group_by(MST) |>
  summarise(n_flights = n_distinct(FLTID), n_rows = n(), .groups = "drop") |>
  arrange(MST) |>
  knitr::kable(caption = "Milestone types present in df_milestones")
```

### HFE Calculation

Calculate HFE indicator for each flight based on milestone pairs.

**HFE formula**: (Great circle distance between milestones / Actual flown distance between milestones) × 100%

```{r}
#| label: calculate-hfe

# Keep only flights with complete trajectories (LAST_HIT within 50 NM of ADES)
complete_flights <- df_milestones |>
  filter(MST == "LAST_HIT", GC_TO_ADES_NM <= 50) |>
  pull(FLTID)

# Calculate HFE for 40 NM milestone pair
hfe_40nm <- df_milestones |>
  filter(FLTID %in% complete_flights) |>
  filter(MST %in% c("FIRST_HIT", "40NM_ADEP", "40NM_ADES", "LAST_HIT")) |>
  group_by(FLTID, ADEP, ADES) |>
  filter(sum(MST %in% c("40NM_ADEP", "40NM_ADES")) == 2) |>  # Both 40NM milestones must exist
  summarise(
    # Takeoff/landing proxies
    takeoff_time = TIME[MST == "FIRST_HIT"][1],
    landing_time = TIME[MST == "LAST_HIT"][1],
    # 40NM milestone points
    start_time = TIME[MST == "40NM_ADEP"][1],
    end_time = TIME[MST == "40NM_ADES"][1],
    start_lat = LAT[MST == "40NM_ADEP"][1],
    start_lon = LON[MST == "40NM_ADEP"][1],
    end_lat = LAT[MST == "40NM_ADES"][1],
    end_lon = LON[MST == "40NM_ADES"][1],
    dist_start = DIST_FLOWN_NM[MST == "40NM_ADEP"][1],
    dist_end = DIST_FLOWN_NM[MST == "40NM_ADES"][1],
    gc_from_adep_start = GC_FROM_ADEP_NM[MST == "40NM_ADEP"][1],
    gc_to_ades_start   = GC_TO_ADES_NM[MST == "40NM_ADEP"][1],
    gc_from_adep_end   = GC_FROM_ADEP_NM[MST == "40NM_ADES"][1],
    gc_to_ades_end     = GC_TO_ADES_NM[MST == "40NM_ADES"][1],
    .groups = "drop"
  ) |>
  mutate(
    # Flown distance along trajectory between milestones (NM)
    flown_dist_nm = dist_end - dist_start,
    # Achieved distance: great-circle between the two milestone points (N→X)
    achieved_dist_nm = great_circle_distance(start_lat, start_lon, end_lat, end_lon),
    hfe_40nm = (achieved_dist_nm / flown_dist_nm) * 100,
    milestone_pair = "40NM"
  ) |>
  filter(flown_dist_nm > 0)  # Remove routes too short for this milestone pair

# Calculate HFE for 100 NM milestone pair
hfe_100nm <- df_milestones |>
  filter(FLTID %in% complete_flights) |>
  filter(MST %in% c("FIRST_HIT", "100NM_ADEP", "100NM_ADES", "LAST_HIT")) |>
  group_by(FLTID, ADEP, ADES) |>
  filter(sum(MST %in% c("100NM_ADEP", "100NM_ADES")) == 2) |>  # Both 100NM milestones must exist
  summarise(
    # Takeoff/landing proxies
    takeoff_time = TIME[MST == "FIRST_HIT"][1],
    landing_time = TIME[MST == "LAST_HIT"][1],
    # 100NM milestone points
    start_time = TIME[MST == "100NM_ADEP"][1],
    end_time = TIME[MST == "100NM_ADES"][1],
    start_lat = LAT[MST == "100NM_ADEP"][1],
    start_lon = LON[MST == "100NM_ADEP"][1],
    end_lat = LAT[MST == "100NM_ADES"][1],
    end_lon = LON[MST == "100NM_ADES"][1],
    dist_start = DIST_FLOWN_NM[MST == "100NM_ADEP"][1],
    dist_end = DIST_FLOWN_NM[MST == "100NM_ADES"][1],
    gc_from_adep_start = GC_FROM_ADEP_NM[MST == "100NM_ADEP"][1],
    gc_to_ades_start   = GC_TO_ADES_NM[MST == "100NM_ADEP"][1],
    gc_from_adep_end   = GC_FROM_ADEP_NM[MST == "100NM_ADES"][1],
    gc_to_ades_end     = GC_TO_ADES_NM[MST == "100NM_ADES"][1],
    .groups = "drop"
  ) |>
  mutate(
    flown_dist_nm = dist_end - dist_start,
    achieved_dist_nm = great_circle_distance(start_lat, start_lon, end_lat, end_lon),
    hfe_100nm = (achieved_dist_nm / flown_dist_nm) * 100,
    milestone_pair = "100NM"
  ) |>
  filter(flown_dist_nm > 0)

# Calculate HFE for V2 milestone pair (40NM ADEP + 100NM ADES, asymmetric)
# V2 = 40NM_ADEP (start) + 100NM_ADES (end) — same milestones as v1/100nm but different pairing
hfe_v2 <- df_milestones |>
  filter(FLTID %in% complete_flights) |>
  filter(MST %in% c("FIRST_HIT", "40NM_ADEP", "100NM_ADES", "LAST_HIT")) |>
  group_by(FLTID, ADEP, ADES) |>
  filter(sum(MST %in% c("40NM_ADEP", "100NM_ADES")) == 2) |>  # Both milestones must exist
  summarise(
    # Takeoff/landing proxies
    takeoff_time = TIME[MST == "FIRST_HIT"][1],
    landing_time = TIME[MST == "LAST_HIT"][1],
    # V2 milestone points: 40NM from ADEP → 100NM from ADES
    start_time = TIME[MST == "40NM_ADEP"][1],
    end_time = TIME[MST == "100NM_ADES"][1],
    start_lat = LAT[MST == "40NM_ADEP"][1],
    start_lon = LON[MST == "40NM_ADEP"][1],
    end_lat = LAT[MST == "100NM_ADES"][1],
    end_lon = LON[MST == "100NM_ADES"][1],
    dist_start = DIST_FLOWN_NM[MST == "40NM_ADEP"][1],
    dist_end = DIST_FLOWN_NM[MST == "100NM_ADES"][1],
    gc_from_adep_start = GC_FROM_ADEP_NM[MST == "40NM_ADEP"][1],
    gc_to_ades_start   = GC_TO_ADES_NM[MST == "40NM_ADEP"][1],
    gc_from_adep_end   = GC_FROM_ADEP_NM[MST == "100NM_ADES"][1],
    gc_to_ades_end     = GC_TO_ADES_NM[MST == "100NM_ADES"][1],
    .groups = "drop"
  ) |>
  mutate(
    flown_dist_nm = dist_end - dist_start,
    achieved_dist_nm = great_circle_distance(start_lat, start_lon, end_lat, end_lon),
    hfe_v2 = (achieved_dist_nm / flown_dist_nm) * 100,
    milestone_pair = "V2 (40+100NM)"
  ) |>
  filter(flown_dist_nm > 0)

# Summary statistics
bind_rows(
  hfe_40nm |> select(FLTID, ADEP, ADES, milestone_pair, hfe = hfe_40nm),
  hfe_100nm |> select(FLTID, ADEP, ADES, milestone_pair, hfe = hfe_100nm),
  hfe_v2 |> select(FLTID, ADEP, ADES, milestone_pair, hfe = hfe_v2)
) |>
  group_by(milestone_pair) |>
  summarise(
    n_flights = n(),
    mean_hfe = mean(hfe, na.rm = TRUE),
    median_hfe = median(hfe, na.rm = TRUE),
    min_hfe = min(hfe, na.rm = TRUE),
    max_hfe = max(hfe, na.rm = TRUE),
    .groups = "drop"
  ) |>
  knitr::kable(
    digits = 2,
    caption = "HFE Summary Statistics by Milestone Pair"
  )
```

## Plotting helpers

```{r}
#| label: plotting-helpers

# Generate a circle of radius_nm NM around (lat0, lon0)
circle_nm <- function(lat0, lon0, radius_nm, n = 180) {
  R_nm <- 3440.065
  angles <- seq(0, 2 * pi, length.out = n + 1)
  lat_r  <- lat0 * pi / 180
  dlat   <- (radius_nm / R_nm) * (180 / pi)
  dlon   <- dlat / cos(lat_r)
  tibble(lon = lon0 + dlon * sin(angles),
         lat = lat0 + dlat * cos(angles))
}

# Generic horizontal trajectory map for any set of flight IDs.
# radius_nm: 40 (V1 symmetric), 100, or c(40, 100) for both circles.
# Shows: base map, actual trajectories, exclusion circles, reference GC line, milestone points.
plot_investigation_horizontal <- function(flt_ids, radius_nm = 40,
                                          title_str = NULL, subtitle_str = NULL) {
  df_traj <- df_trajectories |>
    filter(fid %in% flt_ids) |>
    drop_na(longitude, latitude)

  apt <- df_traj |>
    distinct(adep, ades,
             lon_adep = longitude_adep, lat_adep = latitude_adep,
             lon_ades = longitude_ades, lat_ades = latitude_ades) |>
    slice_head(n = 1)

  # Build one circle per radius per airport
  circles <- purrr::map_dfr(radius_nm, function(r) {
    bind_rows(
      circle_nm(apt$lat_adep, apt$lon_adep, r) |> mutate(airport = "ADEP", r = r),
      circle_nm(apt$lat_ades, apt$lon_ades, r) |> mutate(airport = "ADES", r = r)
    )
  })

  # Milestone labels depend on the first radius supplied (used for reference line)
  r_adep <- radius_nm[1]
  r_ades <- if (length(radius_nm) > 1) radius_nm[2] else radius_nm[1]
  ms_adep <- paste0(r_adep, "NM_ADEP")
  ms_ades <- paste0(r_ades, "NM_ADES")

  ref_segs <- df_milestones |>
    filter(FLTID %in% flt_ids, MST %in% c(ms_adep, ms_ades)) |>
    select(FLTID, MST, LON, LAT) |>
    pivot_wider(names_from = MST, values_from = c(LON, LAT)) |>
    rename(lon_n = !!paste0("LON_", ms_adep), lat_n = !!paste0("LAT_", ms_adep),
           lon_x = !!paste0("LON_", ms_ades), lat_x = !!paste0("LAT_", ms_ades)) |>
    drop_na()

  ms_pts <- ref_segs |>
    pivot_longer(cols = c(lon_n, lat_n, lon_x, lat_x),
                 names_to  = c(".value", "ms"),
                 names_pattern = "(lon|lat)_(n|x)")

  brazil <- rnaturalearth::ne_countries(country = "Brazil",
                                        scale = "medium", returnclass = "sf")

  pad  <- 2
  xlim <- c(min(df_traj$longitude) - pad, max(df_traj$longitude) + pad)
  ylim <- c(min(df_traj$latitude)  - pad, max(df_traj$latitude)  + pad)

  r_label <- paste(radius_nm, collapse = " + ")
  ttl <- title_str    %||% paste0(apt$adep, " → ", apt$ades, " trajectories")
  sub <- subtitle_str %||%
    paste0("Dashed circles = ", r_label, " NM exclusion zone  |  Dotted line = reference GC (N→X)")

  ggplot() +
    geom_sf(data = brazil, fill = "grey95", color = "grey60") +
    geom_path(data = circles,
              aes(x = lon, y = lat, group = interaction(airport, r)),
              color = "#f5a623", linewidth = 0.9, linetype = "dashed") +
    geom_segment(data = ref_segs,
                 aes(x = lon_n, y = lat_n, xend = lon_x, yend = lat_x,
                     color = as.factor(FLTID)),
                 linewidth = 1, linetype = "dotted", alpha = 0.9) +
    geom_path(data = df_traj,
              aes(x = longitude, y = latitude, group = fid, color = as.factor(fid)),
              linewidth = 0.8, alpha = 0.8) +
    geom_point(data = ms_pts,
               aes(x = lon, y = lat, color = as.factor(FLTID)),
               shape = 21, fill = "white", size = 3, stroke = 1.2) +
    geom_point(data = apt, aes(x = lon_adep, y = lat_adep),
               color = "black", size = 4, shape = 17) +
    geom_point(data = apt, aes(x = lon_ades, y = lat_ades),
               color = "black", size = 4, shape = 17) +
    geom_text(data = apt, aes(x = lon_adep, y = lat_adep, label = adep),
              vjust = -1, fontface = "bold") +
    geom_text(data = apt, aes(x = lon_ades, y = lat_ades, label = ades),
              vjust = -1, fontface = "bold") +
    coord_sf(xlim = xlim, ylim = ylim) +
    labs(title = ttl, subtitle = sub,
         x = "Longitude", y = "Latitude", color = "Flight ID") +
    theme_minimal() +
    theme(legend.position = "bottom")
}
```

## Example Visualizations

Visualize sample trajectories for two aerodrome pairs: SBCF-SBKP and SBBR-SBSV.

```{r}
#| label: select-example-pairs

# Select two aerodrome pairs for visualization
example_pairs <- tribble(
  ~adep, ~ades,
  "SBCF", "SBKP",
  "SBBR", "SBSV"
)

# Get flight IDs for these pairs
example_flights <- df_trajectories |>
  inner_join(example_pairs, by = c("adep", "ades")) |>
  group_by(adep, ades) |>
  slice_head(n = 3) |>  # First 3 flights per pair
  distinct(fid) |>
  pull(fid)

# Filter trajectories for examples
df_examples <- df_trajectories |>
  filter(fid %in% example_flights) |>
  drop_na(longitude, latitude, altitude)
```

### Milestone Data Points

The table below shows the extracted milestone positions for the example flights, demonstrating the data points used in HFE calculation:

```{r}
#| label: show-example-milestones

# Filter milestones for example flights only
df_milestones |>
  filter(FLTID %in% example_flights) |>
  arrange(ADEP, ADES, FLTID, TIME) |>
  select(
    FLTID, ADEP, ADES, MST, TIME,
    LAT, LON, ALT,
    `Flown (NM)` = DIST_FLOWN_NM,
    `From ADEP (NM)` = GC_FROM_ADEP_NM,
    `To ADES (NM)` = GC_TO_ADES_NM
  ) |>
  knitr::kable(
    digits = 2,
    caption = "Milestone positions for example flights (SBCF-SBKP and SBBR-SBSV)"
  )
```

### Horizontal (Plan) Views

```{r}
#| label: plot-horizontal-sbcf-sbkp
#| fig-width: 10
#| fig-height: 8

# SBCF - SBKP horizontal view
df_sbcf_sbkp <- df_examples |>
  filter(adep == "SBCF", ades == "SBKP")

# Airport positions
apt_sbcf_sbkp <- df_sbcf_sbkp |>
  distinct(
    adep, ades,
    lon_adep = longitude_adep, lat_adep = latitude_adep,
    lon_ades = longitude_ades, lat_ades = latitude_ades,
    name_adep, name_ades
  )

# 40 NM exclusion circles around ADEP and ADES
circle_adep <- circle_nm(apt_sbcf_sbkp$lat_adep, apt_sbcf_sbkp$lon_adep, 40)
circle_ades <- circle_nm(apt_sbcf_sbkp$lat_ades, apt_sbcf_sbkp$lon_ades, 40)

# Reference segments: GC line between 40NM_ADEP and 40NM_ADES milestones, per flight
ref_segments <- df_milestones |>
  filter(FLTID %in% example_flights, ADEP == "SBCF", ADES == "SBKP") |>
  filter(MST %in% c("40NM_ADEP", "40NM_ADES")) |>
  select(FLTID, MST, LON, LAT) |>
  pivot_wider(names_from = MST, values_from = c(LON, LAT)) |>
  rename(
    lon_n = `LON_40NM_ADEP`, lat_n = `LAT_40NM_ADEP`,
    lon_x = `LON_40NM_ADES`, lat_x = `LAT_40NM_ADES`
  ) |>
  drop_na()

# Milestone points (N and X) for labelling
ms_points <- ref_segments |>
  pivot_longer(
    cols = c(lon_n, lat_n, lon_x, lat_x),
    names_to = c(".value", "ms"),
    names_pattern = "(lon|lat)_(n|x)"
  )

ggplot(df_sbcf_sbkp) +
  # 40 NM circles (exclusion zones)
  geom_path(data = circle_adep, aes(x = lon, y = lat),
            color = "#f5a623", linewidth = 0.8, linetype = "dashed") +
  geom_path(data = circle_ades, aes(x = lon, y = lat),
            color = "#f5a623", linewidth = 0.8, linetype = "dashed") +
  # Reference line (GC between milestones N → X) per flight
  geom_segment(data = ref_segments,
               aes(x = lon_n, y = lat_n, xend = lon_x, yend = lat_x,
                   color = as.factor(FLTID)),
               linewidth = 1, linetype = "dotted", alpha = 0.9) +
  # Actual trajectories
  geom_path(aes(x = longitude, y = latitude, group = fid, color = as.factor(fid)),
            linewidth = 0.8, alpha = 0.7) +
  # Milestone points N and X
  geom_point(data = ms_points,
             aes(x = lon, y = lat, color = as.factor(FLTID)),
             shape = 21, fill = "white", size = 3, stroke = 1.2) +
  # Airport markers
  geom_point(data = apt_sbcf_sbkp,
             aes(x = lon_adep, y = lat_adep),
             color = "black", size = 4, shape = 17) +
  geom_point(data = apt_sbcf_sbkp,
             aes(x = lon_ades, y = lat_ades),
             color = "black", size = 4, shape = 17) +
  geom_text(data = apt_sbcf_sbkp,
            aes(x = lon_adep, y = lat_adep, label = adep),
            vjust = -1, fontface = "bold") +
  geom_text(data = apt_sbcf_sbkp,
            aes(x = lon_ades, y = lat_ades, label = ades),
            vjust = -1, fontface = "bold") +
  coord_fixed() +
  labs(
    title = "Horizontal View: SBCF → SBKP",
    subtitle = sprintf(
      "%s → %s  |  Dashed circles = 40 NM exclusion zone  |  Dotted line = reference (GC N→X)",
      apt_sbcf_sbkp$name_adep, apt_sbcf_sbkp$name_ades
    ),
    x = "Longitude", y = "Latitude",
    color = "Flight ID"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r}
#| label: plot-horizontal-sbbr-sbsv
#| fig-width: 10
#| fig-height: 8

# SBBR - SBSV horizontal view
df_sbbr_sbsv <- df_examples |>
  filter(adep == "SBBR", ades == "SBSV")

# Airport positions
apt_sbbr_sbsv <- df_sbbr_sbsv |>
  distinct(
    adep, ades,
    lon_adep = longitude_adep, lat_adep = latitude_adep,
    lon_ades = longitude_ades, lat_ades = latitude_ades,
    name_adep, name_ades
  )

ggplot(df_sbbr_sbsv) +
  geom_path(aes(x = longitude, y = latitude, group = fid, color = as.factor(fid)),
            linewidth = 0.8, alpha = 0.7) +
  geom_point(data = apt_sbbr_sbsv,
             aes(x = lon_adep, y = lat_adep),
             color = "green", size = 4, shape = 17) +
  geom_point(data = apt_sbbr_sbsv,
             aes(x = lon_ades, y = lat_ades),
             color = "red", size = 4, shape = 17) +
  geom_text(data = apt_sbbr_sbsv,
            aes(x = lon_adep, y = lat_adep, label = adep),
            vjust = -1, color = "green", fontface = "bold") +
  geom_text(data = apt_sbbr_sbsv,
            aes(x = lon_ades, y = lat_ades, label = ades),
            vjust = -1, color = "red", fontface = "bold") +
  coord_fixed() +
  labs(
    title = "Horizontal View: SBBR → SBSV",
    subtitle = sprintf("%s → %s", apt_sbbr_sbsv$name_adep, apt_sbbr_sbsv$name_ades),
    x = "Longitude", y = "Latitude",
    color = "Flight ID"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Vertical Profiles

```{r}
#| label: plot-vertical-sbcf-sbkp
#| fig-width: 10
#| fig-height: 6

# Use trrrj plotting function
df_sbcf_sbkp |>
  cumulative_distance() |>
  plot_flight_vertical_distance() +
  labs(
    title = "Vertical Profile: SBCF → SBKP",
    subtitle = sprintf("%s → %s", apt_sbcf_sbkp$name_adep, apt_sbcf_sbkp$name_ades)
  )
```

```{r}
#| label: plot-vertical-sbbr-sbsv
#| fig-width: 10
#| fig-height: 6

# Use trrrj plotting function
df_sbbr_sbsv |>
  cumulative_distance() |>
  plot_flight_vertical_distance() +
  labs(
    title = "Vertical Profile: SBBR → SBSV",
    subtitle = sprintf("%s → %s", apt_sbbr_sbsv$name_adep, apt_sbbr_sbsv$name_ades)
  )
```

## Comparison with Brazil HFE Data

Compare our calculated milestones and HFE values against Brazil's reference dataset.

**Note on time matching**:

- **Our Time**: FIRST_HIT timestamp (first surveillance point, proxy for takeoff)
- **BRA Time**: Timestamp from Brazil HFE dataset `time` field
- **Matching**: Flights matched by ADEP/ADES pair and time proximity (within 1 hour)

```{r}
#| label: match-brazil-flights

# Try to match flights between our extract and Brazil HFE
# Match criteria: same ADEP, ADES, and similar time (within 1 hour)

flight_comparison <- hfe_40nm |>
  mutate(
    adep = ADEP,
    ades = ADES,
    our_time = takeoff_time,  # Use FIRST_HIT as takeoff proxy
    our_hfe_40nm = hfe_40nm,
    our_achieved_dist_nm = achieved_dist_nm,
    our_flown_dist_nm = flown_dist_nm
  ) |>
  left_join(
    hfe_v2 |> select(FLTID, our_hfe_v2 = hfe_v2, our_flown_dist_v2_nm = flown_dist_nm),
    by = "FLTID"
  ) |>
  left_join(
    bra_hfe_sample |>
      mutate(
        bra_time = lubridate::as_datetime(time),
        bra_hfe_v1 = kpi05_v1,
        bra_hfe_v2 = kpi05_v2,
        bra_dist_flown_v1_nm = dist_flown_v1,
        bra_dist_flown_v2_nm = dist_flown_v2
      ) |>
      select(
        adep = plan_dep,
        ades = plan_arr,
        bra_time,
        bra_callsign = id_icao,
        bra_hfe_v1,
        bra_hfe_v2,
        bra_dist_flown_v1_nm = dist_flown_v1,
        bra_dist_flown_v2_nm = dist_flown_v2
      ),
    by = c("adep", "ades"),
    relationship = "many-to-many"
  ) |>
  # Match by time proximity (within 1 hour)
  mutate(time_diff_min = abs(as.numeric(difftime(our_time, bra_time, units = "mins")))) |>
  group_by(FLTID) |>
  slice_min(time_diff_min, n = 1) |>
  ungroup() |>
  filter(time_diff_min <= 15) |>  # Keep only matches within 15 minutes
  mutate(
    dist_diff_nm = abs(our_flown_dist_nm - bra_dist_flown_v1_nm),
    valid = dist_diff_nm <= 50  # Flag: TRUE = trajectory consistent with BRA reference
  )

flight_comparison |>
  mutate(
    # kpi05_v1/v2 is additional NM flown — convert to HFE percentage
    bra_hfe_v1_pct = (1 - bra_hfe_v1 / bra_dist_flown_v1_nm) * 100,
    bra_hfe_v2_pct = (1 - bra_hfe_v2 / bra_dist_flown_v2_nm) * 100
  ) |>
  select(
    FLTID, ADEP, ADES,
    `Our Takeoff` = our_time,
    `BRA Time` = bra_time,
    `Δ Time (min)` = time_diff_min,
    `Valid` = valid,
    `Dist Diff (NM)` = dist_diff_nm,
    `Our HFE v1 (%)` = our_hfe_40nm,
    `Our HFE v2 (%)` = our_hfe_v2,
    `BRA HFE v1 (%)` = bra_hfe_v1_pct,
    `BRA HFE v2 (%)` = bra_hfe_v2_pct,
    `Our Flown v1 (NM)` = our_flown_dist_nm,
    `Our Flown v2 (NM)` = our_flown_dist_v2_nm,
    `BRA Dist v1 (NM)` = bra_dist_flown_v1_nm,
    `BRA Dist v2 (NM)` = bra_dist_flown_v2_nm
  ) |>
  head(15) |>
  knitr::kable(
    digits = 2,
    caption = "Flight matching: Our calculations vs Brazil HFE reference"
  )
```

```{r}
#| label: comparison-stats

# Summary statistics of differences
if (nrow(flight_comparison) > 0) {
  flight_comparison |>
    mutate(
      bra_hfe_v1_pct = (1 - bra_hfe_v1 / bra_dist_flown_v1_nm) * 100,
      bra_hfe_v2_pct = (1 - bra_hfe_v2 / bra_dist_flown_v2_nm) * 100
    ) |>
    group_by(valid) |>
    summarise(
      n_flights = n(),
      mean_hfe_diff_v1 = mean(abs(our_hfe_40nm - bra_hfe_v1_pct), na.rm = TRUE),
      mean_hfe_diff_v2 = mean(abs(our_hfe_v2 - bra_hfe_v2_pct), na.rm = TRUE),
      mean_dist_diff_v1 = mean(abs(our_flown_dist_nm - bra_dist_flown_v1_nm), na.rm = TRUE),
      mean_dist_diff_v2 = mean(abs(our_flown_dist_v2_nm - bra_dist_flown_v2_nm), na.rm = TRUE),
      .groups = "drop"
    ) |>
    knitr::kable(
      digits = 2,
      caption = "Comparison stats by validity flag (threshold: 50 NM distance difference)"
    )
} else {
  cat("No flights matched between datasets")
}
```

```{r}
#| label: comparison-by-pair

# Breakdown by aerodrome pair — valid flights only
flight_comparison |>
  filter(valid == TRUE) |>
  mutate(
    bra_hfe_v1_pct = (1 - bra_hfe_v1 / bra_dist_flown_v1_nm) * 100,
    bra_hfe_v2_pct = (1 - bra_hfe_v2 / bra_dist_flown_v2_nm) * 100,
    hfe_diff_v1 = our_hfe_40nm - bra_hfe_v1_pct,
    hfe_diff_v2 = our_hfe_v2 - bra_hfe_v2_pct,
    dist_diff_v1 = our_flown_dist_nm - bra_dist_flown_v1_nm,
    dist_diff_v2 = our_flown_dist_v2_nm - bra_dist_flown_v2_nm
  ) |>
  group_by(ADEP, ADES) |>
  summarise(
    n = n(),
    mean_our_hfe_v1  = mean(our_hfe_40nm,   na.rm = TRUE),
    mean_bra_hfe_v1  = mean(bra_hfe_v1_pct, na.rm = TRUE),
    mean_diff_v1     = mean(hfe_diff_v1,     na.rm = TRUE),
    mean_our_hfe_v2  = mean(our_hfe_v2,      na.rm = TRUE),
    mean_bra_hfe_v2  = mean(bra_hfe_v2_pct,  na.rm = TRUE),
    mean_diff_v2     = mean(hfe_diff_v2,      na.rm = TRUE),
    mean_dist_diff_v1 = mean(dist_diff_v1,   na.rm = TRUE),
    mean_dist_diff_v2 = mean(dist_diff_v2,   na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(abs(mean_diff_v2))) |>
  knitr::kable(
    digits = 2,
    caption = "HFE comparison by aerodrome pair (valid flights only) — positive diff = our HFE higher than BRA"
  )
```

## Investigation: SBSV ↔ SBGR

SBSV→SBGR shows a systematic −7.19% difference (our HFE 91.64% vs BRA 98.83%). This section investigates the cause.

```{r}
#| label: investigate-sbsv-sbgr
#| fig-width: 10
#| fig-height: 8

# Step 1: Flight-level detail for SBSV→SBGR
flight_comparison |>
  filter(valid == TRUE, ADEP == "SBSV", ADES == "SBGR") |>
  mutate(
    bra_hfe_v1_pct  = (1 - bra_hfe_v1 / bra_dist_flown_v1_nm) * 100,
    # BRA achieved GC = flown * (hfe/100)
    bra_achieved_nm = bra_dist_flown_v1_nm * (bra_hfe_v1_pct / 100),
    hfe_diff        = our_hfe_40nm - bra_hfe_v1_pct
  ) |>
  select(
    FLTID,
    `Takeoff`          = our_time,
    `Our Flown (NM)`   = our_flown_dist_nm,
    `BRA Flown (NM)`   = bra_dist_flown_v1_nm,
    `Our Achieved (NM)`= our_achieved_dist_nm,
    `BRA Achieved (NM)`= bra_achieved_nm,
    `Our HFE v1 (%)`   = our_hfe_40nm,
    `BRA HFE v1 (%)`   = bra_hfe_v1_pct,
    `Diff (%)`         = hfe_diff
  ) |>
  arrange(`Diff (%)`) |>
  knitr::kable(digits = 2, caption = "SBSV→SBGR: flight-by-flight comparison")
```

```{r}
#| label: investigate-sbsv-sbgr-milestones

# Step 2: Check milestone positions for SBSV→SBGR flights
sbsv_sbgr_fltids <- flight_comparison |>
  filter(valid == TRUE, ADEP == "SBSV", ADES == "SBGR") |>
  pull(FLTID)

df_milestones |>
  filter(FLTID %in% sbsv_sbgr_fltids) |>
  filter(MST %in% c("FIRST_HIT", "40NM_ADEP", "40NM_ADES", "LAST_HIT")) |>
  select(FLTID, MST, TIME,
         `Flown (NM)` = DIST_FLOWN_NM,
         `GC from ADEP (NM)` = GC_FROM_ADEP_NM,
         `GC to ADES (NM)` = GC_TO_ADES_NM) |>
  arrange(FLTID, TIME) |>
  knitr::kable(digits = 2, caption = "SBSV→SBGR: milestone positions")
```

```{r}
#| label: plot-sbsv-sbgr
#| fig-width: 11
#| fig-height: 9

# Step 3: Trajectory map for SBSV→SBGR with reference lines and exclusion circles
plot_investigation_horizontal(
  sbsv_sbgr_fltids,
  title_str    = "SBSV → SBGR trajectories",
  subtitle_str = "Salvador → São Paulo (Guarulhos)  |  Dashed circles = 40 NM  |  Dotted = reference GC"
)
```

```{r}
#| label: investigate-sbsv-sbgr-gc

# Step 4: How far does gc_to_ades actually go for these flights?
# Check if 40NM_ADES milestone is far from ADES (route longer than expected)
df_with_gc |>
  filter(fid %in% sbsv_sbgr_fltids) |>
  group_by(fid) |>
  summarise(
    total_flown_nm = max(cum_dist, na.rm = TRUE),
    gc_adep_to_ades = first(gc_to_ades),      # GC at first point ≈ total GC distance
    max_gc_from_adep = max(gc_from_adep, na.rm = TRUE),
    min_gc_to_ades = min(gc_to_ades, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    extra_flown_nm = total_flown_nm - gc_adep_to_ades,
    hfe_approx = (gc_adep_to_ades / total_flown_nm) * 100
  ) |>
  knitr::kable(digits = 2, caption = "SBSV→SBGR: raw trajectory stats (gate-to-gate)")
```
