---
title: "Flight Matching Diagnostic & Improved Matching"
author: "PRU Analysis"
date: today
format:
  html:
    code-fold: true
    toc: true
execute:
  warning: false
---

## Overview

This document investigates why the initial flight matching failed and implements improved matching strategies.

**Matching Challenge**: Link flights from our radar extract to Brazil's HFE reference dataset.

## Setup

```{r}
#| label: setup
#| message: false

library(conflicted)
library(tidyverse)
library(arrow)
library(lubridate)

conflict_prefer("filter", "dplyr")
conflicts_prefer(dplyr::lag)

# Load local helper functions
source("R/sf_helpers.R")

# Path to radar_brazil utilities
radar_utils_path <- "../radar_brazil"
```

## Load Data

```{r}
#| label: load-data

# Our extracted trajectories
df_trajectories <- read_parquet("data/sample_trajectories_2025_12_10.parquet")

# Brazil HFE reference data
bra_hfe_sample <- read_delim(
  "data/brazil_hfe_2025_12_10.csv",
  delim = ",",
  trim_ws = TRUE,
  show_col_types = FALSE
)

# Get our flight summary
our_flights <- df_trajectories |>
  group_by(fid) |>
  summarise(
    adep = first(adep),
    ades = first(ades),
    callsign = first(callsign),
    first_time = min(timestamp),
    last_time = max(timestamp),
    n_positions = n(),
    .groups = "drop"
  ) |>
  arrange(adep, ades, first_time)

# Clean Brazil HFE data
bra_flights <- bra_hfe_sample |>
  mutate(
    bra_time = as_datetime(time),
    bra_date = as_date(bra_time),
    bra_callsign = id_icao
  ) |>
  rename(
    adep = plan_dep,
    ades = plan_arr
  ) |>
  select(
    adep, ades, bra_callsign, bra_time, bra_date,
    bra_hfe_v1 = kpi05_v1,
    bra_hfe_v2 = kpi05_v2,
    bra_dist_flown_v1 = dist_flown_v1,
    bra_dist_flown_v2 = dist_flown_v2,
    bra_dist_achieved_v1 = dist_achieved_v1,
    bra_dist_achieved_v2 = dist_achieved_v2,
    bra_equip = equip
  ) |>
  arrange(adep, ades, bra_time)
```

## Diagnostic 1: Data Coverage Comparison

Compare flight counts per route:

```{r}
#| label: coverage-comparison

coverage_comparison <- our_flights |>
  count(adep, ades, name = "our_count") |>
  full_join(
    bra_flights |> count(adep, ades, name = "bra_count"),
    by = c("adep", "ades")
  ) |>
  replace_na(list(our_count = 0, bra_count = 0)) |>
  mutate(
    difference = bra_count - our_count,
    pct_coverage = round(100 * our_count / bra_count, 1)
  ) |>
  arrange(adep, ades)

coverage_comparison |>
  knitr::kable(
    caption = "Flight count comparison by route",
    col.names = c("ADEP", "ADES", "Our Extract", "Brazil HFE", "Difference", "% Coverage")
  )
```

**Key Observations**:

```{r}
#| label: coverage-stats

list(
  total_our = sum(coverage_comparison$our_count),
  total_brazil = sum(coverage_comparison$bra_count),
  overall_coverage_pct = round(100 * sum(coverage_comparison$our_count) / sum(coverage_comparison$bra_count), 1),
  routes_with_perfect_match = sum(coverage_comparison$difference == 0),
  routes_missing_flights = sum(coverage_comparison$difference > 0)
)
```

## Diagnostic 2: Time Field Analysis

Investigate time differences between datasets:

```{r}
#| label: time-analysis

# For routes with flights in both datasets, examine time distributions
time_comparison <- our_flights |>
  select(adep, ades, our_time = first_time) |>
  inner_join(
    bra_flights |> select(adep, ades, bra_time),
    by = c("adep", "ades"),
    relationship = "many-to-many"
  ) |>
  mutate(
    time_diff_min = as.numeric(difftime(our_time, bra_time, units = "mins")),
    abs_time_diff_min = abs(time_diff_min)
  )

# Summary statistics
time_comparison |>
  summarise(
    n_comparisons = n(),
    mean_abs_diff = mean(abs_time_diff_min),
    median_abs_diff = median(abs_time_diff_min),
    min_abs_diff = min(abs_time_diff_min),
    max_abs_diff = max(abs_time_diff_min),
    within_5min = sum(abs_time_diff_min <= 5),
    within_15min = sum(abs_time_diff_min <= 15),
    within_30min = sum(abs_time_diff_min <= 30),
    within_60min = sum(abs_time_diff_min <= 60)
  ) |>
  knitr::kable(
    digits = 1,
    caption = "Time difference statistics (all pairwise comparisons)"
  )
```

### Time Difference Distribution

```{r}
#| label: plot-time-diff
#| fig-width: 10
#| fig-height: 6

ggplot(time_comparison, aes(x = abs_time_diff_min)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = c(5, 15, 30, 60),
             linetype = "dashed", color = "red", linewidth = 0.8) +
  scale_x_continuous(breaks = c(0, 5, 15, 30, 60, 120, 180, 240, 300)) +
  labs(
    title = "Distribution of Time Differences (Absolute)",
    subtitle = "All pairwise comparisons between our extract and Brazil HFE",
    x = "Absolute Time Difference (minutes)",
    y = "Count",
    caption = "Red lines: 5, 15, 30, 60 minute thresholds"
  ) +
  theme_minimal()
```

## Diagnostic 3: Callsign Matching

Check if callsigns match between datasets:

```{r}
#| label: callsign-analysis

# Join on route and check callsign overlap
callsign_comparison <- our_flights |>
  select(fid, adep, ades, our_callsign = callsign, our_time = first_time) |>
  inner_join(
    bra_flights |> select(adep, ades, bra_callsign, bra_time),
    by = c("adep", "ades"),
    relationship = "many-to-many"
  ) |>
  mutate(
    callsign_match = our_callsign == bra_callsign,
    time_diff_min = abs(as.numeric(difftime(our_time, bra_time, units = "mins")))
  )

# How many have matching callsigns?
callsign_comparison |>
  summarise(
    total_comparisons = n(),
    callsign_matches = sum(callsign_match, na.rm = TRUE),
    pct_callsign_match = round(100 * callsign_matches / total_comparisons, 1),
    matches_within_15min = sum(callsign_match & time_diff_min <= 15, na.rm = TRUE),
    matches_within_60min = sum(callsign_match & time_diff_min <= 60, na.rm = TRUE)
  ) |>
  knitr::kable(
    caption = "Callsign matching analysis"
  )
```

### Sample Callsigns

```{r}
#| label: sample-callsigns

# Show sample of callsigns from each dataset
list(
  our_callsigns_sample = head(sort(unique(our_flights$callsign)), 15),
  bra_callsigns_sample = head(sort(unique(bra_flights$bra_callsign)), 15)
)
```

## Strategy 1: Nearest Neighbor Matching (Time-Based)

Match each of our flights to the nearest Brazil HFE flight by time (within threshold).

**Strategy**: For each route pair, find the Brazil flight closest in time to our flight.

### Example: When Time Matching Works

```{r}
#| label: time-matching-success-example

# Find a good time-match example (same route, close time, different callsigns)
time_match_example <- our_flights |>
  inner_join(
    bra_flights,
    by = c("adep", "ades"),
    relationship = "many-to-many"
  ) |>
  mutate(
    time_diff_min = abs(as.numeric(difftime(first_time, bra_time, units = "mins"))),
    callsign_match = callsign == bra_callsign
  ) |>
  filter(!callsign_match, time_diff_min < 2) |>
  arrange(time_diff_min) |>
  head(1)

if (nrow(time_match_example) > 0) {
  time_match_example |>
    select(
      fid, adep, ades,
      our_callsign = callsign,
      bra_callsign,
      our_time = first_time,
      bra_time,
      time_diff_min
    ) |>
    knitr::kable(
      digits = 2,
      caption = "✅ SUCCESS: Time matching works when callsigns differ but timing is tight"
    )
} else {
  cat("Most flights match by callsign, so time-only matches are rare.\n")
}
```

### Counter-Example: When Time Matching Fails

```{r}
#| label: time-matching-failure-example

# Find cases where multiple flights on same route within short time window
time_ambiguity <- our_flights |>
  inner_join(
    bra_flights,
    by = c("adep", "ades"),
    relationship = "many-to-many"
  ) |>
  mutate(
    time_diff_min = abs(as.numeric(difftime(first_time, bra_time, units = "mins")))
  ) |>
  group_by(fid) |>
  filter(n() > 1, time_diff_min <= 30) |>
  slice_min(time_diff_min, n = 3) |>
  ungroup() |>
  arrange(fid, time_diff_min) |>
  head(6)

if (nrow(time_ambiguity) > 0) {
  time_ambiguity |>
    select(
      fid, adep, ades,
      our_callsign = callsign,
      bra_callsign,
      time_diff_min
    ) |>
    knitr::kable(
      digits = 2,
      caption = "❌ FAILURE: Time matching alone creates ambiguity - which Brazil flight is correct?"
    )
} else {
  cat("Time matching has clear winners in this dataset.\n")
}
```

**Problem with time-only matching**: On busy routes (e.g., SBCF → SBSP with 30 flights/day), multiple flights depart within minutes. Time proximity alone cannot distinguish the correct match.

**Solution**: Need additional discriminator → **Callsign matching** (Strategy 2)

### Implementation

```{r}
#| label: nn-time-matching

match_nearest_by_time <- function(our_df, bra_df, max_time_diff_min = 30) {

  matched <- our_df |>
    left_join(
      bra_df,
      by = c("adep", "ades"),
      relationship = "many-to-many"
    ) |>
    mutate(
      time_diff_min = abs(as.numeric(difftime(first_time, bra_time, units = "mins")))
    ) |>
    # For each of our flights, find closest Brazil flight
    group_by(fid) |>
    slice_min(time_diff_min, n = 1, with_ties = FALSE) |>
    ungroup() |>
    # Only keep if within threshold
    filter(time_diff_min <= max_time_diff_min)

  return(matched)
}

# Try with different time thresholds
thresholds <- c(5, 15, 30, 60)

threshold_results <- map_dfr(thresholds, function(thresh) {
  matched <- match_nearest_by_time(our_flights, bra_flights, max_time_diff_min = thresh)

  tibble(
    threshold_min = thresh,
    n_matched = nrow(matched),
    pct_matched = round(100 * nrow(matched) / nrow(our_flights), 1),
    mean_time_diff = round(mean(matched$time_diff_min), 1),
    median_time_diff = round(median(matched$time_diff_min), 1)
  )
})

threshold_results |>
  knitr::kable(
    caption = "Nearest neighbor time matching results",
    col.names = c("Threshold (min)", "Matched", "% Matched", "Mean Δt", "Median Δt")
  )
```

### Apply Best Time Matching (15 min threshold)

```{r}
#| label: apply-time-matching

matched_time <- match_nearest_by_time(our_flights, bra_flights, max_time_diff_min = 15)

cat(sprintf("Matched %d out of %d flights (%.1f%%) using 15-minute time threshold\n",
            nrow(matched_time), nrow(our_flights),
            100 * nrow(matched_time) / nrow(our_flights)))
```

## Strategy 2: Callsign + Route Matching

Match by callsign and route (ignore time initially).

**Strategy**: Exact match on ADEP + ADES + Callsign, then pick closest by time if multiple candidates.

### Example: When Callsign Matching Works

```{r}
#| label: callsign-matching-success-example

# Find perfect callsign matches
callsign_perfect <- our_flights |>
  inner_join(
    bra_flights,
    by = c("adep", "ades", "callsign" = "bra_callsign")
  ) |>
  mutate(
    time_diff_min = abs(as.numeric(difftime(first_time, bra_time, units = "mins")))
  ) |>
  group_by(fid) |>
  slice_min(time_diff_min, n = 1) |>
  ungroup() |>
  arrange(time_diff_min) |>
  head(3)

callsign_perfect |>
  select(
    fid, adep, ades,
    callsign,
    our_time = first_time,
    bra_time,
    time_diff_min,
    n_positions
  ) |>
  knitr::kable(
    digits = 2,
    caption = "✅ SUCCESS: Callsign matching gives unambiguous identification"
  )
```

**Why this works**: Callsigns are unique per flight on a given route at a given time. Even if time differs slightly, the callsign resolves ambiguity.

### Counter-Example: When Callsign Matching Fails

```{r}
#| label: callsign-matching-failure-example

# Find flights in our data that have NO callsign match in Brazil data
our_callsigns <- unique(our_flights$callsign)
bra_callsigns <- unique(bra_flights$bra_callsign)

callsign_orphans <- our_flights |>
  filter(!callsign %in% bra_callsigns) |>
  arrange(adep, ades, first_time) |>
  head(5)

if (nrow(callsign_orphans) > 0) {
  callsign_orphans |>
    select(
      fid, adep, ades, callsign, first_time, n_positions
    ) |>
    knitr::kable(
      caption = "❌ FAILURE: These callsigns don't appear in Brazil dataset at all"
    )

  cat(sprintf("\n%d flights have callsigns not found in Brazil data.\n",
              sum(!our_flights$callsign %in% bra_callsigns)))
} else {
  cat("All our callsigns have matches in Brazil data.\n")
}
```

**Problem**: Some flights use callsigns that don't exist in Brazil's dataset. Possible reasons:
- Brazil uses different callsign format (e.g., flight number vs radio callsign)
- Flights filtered out by Brazil's quality criteria before callsign recorded
- Radar/ATC system differences in callsign capture

**Solution**: For unmatched callsigns, fall back to time-based matching → **Hybrid approach** (Strategy 3)

### Implementation

```{r}
#| label: callsign-matching

matched_callsign <- our_flights |>
  inner_join(
    bra_flights,
    by = c("adep", "ades", "callsign" = "bra_callsign")
  ) |>
  mutate(
    time_diff_min = abs(as.numeric(difftime(first_time, bra_time, units = "mins")))
  ) |>
  # If multiple matches, take closest by time
  group_by(fid) |>
  slice_min(time_diff_min, n = 1, with_ties = FALSE) |>
  ungroup()

cat(sprintf("Matched %d out of %d flights (%.1f%%) using callsign + route\n",
            nrow(matched_callsign), nrow(our_flights),
            100 * nrow(matched_callsign) / nrow(our_flights)))

# Time difference for callsign matches
matched_callsign |>
  summarise(
    n = n(),
    mean_time_diff = mean(time_diff_min),
    median_time_diff = median(time_diff_min),
    max_time_diff = max(time_diff_min)
  ) |>
  knitr::kable(
    digits = 1,
    caption = "Time differences for callsign matches"
  )
```

## Strategy 3: Hybrid Matching

**Strategy**: Combine the strengths of both approaches:
1. **First pass**: Match by callsign (high confidence, unambiguous)
2. **Second pass**: For remaining flights, match by time proximity (15 min window)

This gives us:
- **Precision** from callsign matching (eliminates false positives)
- **Coverage** from time matching (catches callsign mismatches)

### Why Hybrid is Needed: Concrete Example

```{r}
#| label: hybrid-motivation-example

# Show a flight that WOULD match by time but WRONG callsign
# Then show it correctly matched by hybrid approach

# Example: Flight that needs time-based fallback
needs_time_fallback <- our_flights |>
  filter(!callsign %in% bra_callsigns) |>
  inner_join(
    bra_flights,
    by = c("adep", "ades"),
    relationship = "many-to-many"
  ) |>
  mutate(
    time_diff_min = abs(as.numeric(difftime(first_time, bra_time, units = "mins")))
  ) |>
  group_by(fid) |>
  slice_min(time_diff_min, n = 1) |>
  ungroup() |>
  filter(time_diff_min <= 15) |>
  head(2)

if (nrow(needs_time_fallback) > 0) {
  needs_time_fallback |>
    select(
      fid, adep, ades,
      our_callsign = callsign,
      bra_callsign,
      time_diff_min
    ) |>
    knitr::kable(
      digits = 2,
      caption = "Hybrid Success: These flights need time-based fallback (callsign mismatch)"
    )
} else {
  cat("Most flights in this sample match by callsign.\n")
}
```

**Interpretation**: Without the hybrid approach, we'd lose these flights. They're genuine matches (same route, same time) but with different callsign identifiers.

### Implementation

```{r}
#| label: hybrid-matching

# Step 1: Callsign matches
matched_by_callsign <- our_flights |>
  inner_join(
    bra_flights,
    by = c("adep", "ades", "callsign" = "bra_callsign")
  ) |>
  mutate(
    time_diff_min = abs(as.numeric(difftime(first_time, bra_time, units = "mins"))),
    match_method = "callsign"
  ) |>
  group_by(fid) |>
  slice_min(time_diff_min, n = 1, with_ties = FALSE) |>
  ungroup()

# Step 2: For unmatched flights, try time-based matching
unmatched_fids <- setdiff(our_flights$fid, matched_by_callsign$fid)

our_unmatched <- our_flights |> filter(fid %in% unmatched_fids)

matched_by_time <- match_nearest_by_time(our_unmatched, bra_flights, max_time_diff_min = 15) |>
  mutate(match_method = "time")

# Combine
matched_hybrid <- bind_rows(
  matched_by_callsign,
  matched_by_time
)

# Summary
matched_hybrid |>
  count(match_method) |>
  mutate(pct = round(100 * n / nrow(our_flights), 1)) |>
  knitr::kable(
    caption = "Hybrid matching results",
    col.names = c("Method", "Count", "% of Total")
  )

cat(sprintf("\nTotal matched: %d out of %d flights (%.1f%%)\n",
            nrow(matched_hybrid), nrow(our_flights),
            100 * nrow(matched_hybrid) / nrow(our_flights)))
```

## Matched Flights Summary

```{r}
#| label: matched-summary

matched_hybrid |>
  group_by(adep, ades) |>
  summarise(
    n_matched = n(),
    callsign_matches = sum(match_method == "callsign"),
    time_matches = sum(match_method == "time"),
    mean_time_diff = round(mean(time_diff_min), 1),
    .groups = "drop"
  ) |>
  arrange(desc(n_matched)) |>
  knitr::kable(
    caption = "Matched flights by route",
    col.names = c("ADEP", "ADES", "Matched", "By Callsign", "By Time", "Mean Δt (min)")
  )
```

## Unmatched Flights Analysis

Investigate why some flights didn't match:

```{r}
#| label: unmatched-analysis

unmatched_our <- our_flights |>
  filter(!fid %in% matched_hybrid$fid) |>
  arrange(adep, ades, first_time)

cat(sprintf("Unmatched flights: %d (%.1f%%)\n",
            nrow(unmatched_our),
            100 * nrow(unmatched_our) / nrow(our_flights)))

if (nrow(unmatched_our) > 0) {
  unmatched_our |>
    count(adep, ades) |>
    arrange(desc(n)) |>
    knitr::kable(
      caption = "Unmatched flights by route",
      col.names = c("ADEP", "ADES", "Count")
    )

  # Show sample of unmatched flights
  cat("\nSample of unmatched flights:\n")
  unmatched_our |>
    select(fid, adep, ades, callsign, first_time, last_time, n_positions) |>
    head(10) |>
    knitr::kable()
}
```

## Export Matched Dataset

```{r}
#| label: export-matched

# Create comprehensive matched dataset
matched_export <- matched_hybrid |>
  select(
    fid,
    adep, ades,
    our_callsign = callsign,
    bra_callsign,
    our_first_time = first_time,
    our_last_time = last_time,
    bra_time,
    time_diff_min,
    match_method,
    n_positions,
    bra_hfe_v1,
    bra_hfe_v2,
    bra_dist_flown_v1,
    bra_dist_flown_v2,
    bra_dist_achieved_v1,
    bra_dist_achieved_v2,
    bra_equip
  )

write_csv(matched_export, "data/matched_flights.csv")

cat(sprintf("Exported %d matched flights to data/matched_flights.csv\n", nrow(matched_export)))
```

## Validation Comparison Preview

Quick preview of HFE comparison (detailed analysis in next document):

```{r}
#| label: hfe-preview
#| eval: false

# Load our HFE calculations (from milestone extraction)
# This will be done in the next analysis document
# For now, just show the structure

matched_export |>
  select(
    fid, adep, ades,
    match_method,
    time_diff_min,
    bra_hfe_v1,
    bra_hfe_v2,
    bra_dist_flown_v1,
    bra_dist_flown_v2
  ) |>
  head(15) |>
  knitr::kable(
    digits = 2,
    caption = "Preview: Matched flights ready for HFE comparison"
  )
```

## Deep Dive: Unmatched Flights Investigation

Let's investigate the 32 flights that didn't match to understand why:

```{r}
#| label: unmatched-deep-dive

# Get Brazil flights that weren't matched
matched_bra_ids <- matched_hybrid |>
  select(adep, ades, bra_callsign, bra_time) |>
  mutate(bra_id = paste(adep, ades, bra_callsign, bra_time, sep = "_"))

bra_unmatched <- bra_flights |>
  mutate(bra_id = paste(adep, ades, bra_callsign, bra_time, sep = "_")) |>
  filter(!bra_id %in% matched_bra_ids$bra_id)

cat(sprintf("Our unmatched flights: %d\n", nrow(unmatched_our)))
cat(sprintf("Brazil unmatched flights: %d\n", nrow(bra_unmatched)))
```

### Unmatched Flight Characteristics

```{r}
#| label: unmatched-characteristics

if (nrow(unmatched_our) > 0) {
  # Our unmatched flights by route
  unmatched_our_summary <- unmatched_our |>
    group_by(adep, ades) |>
    summarise(
      n_flights = n(),
      avg_positions = round(mean(n_positions)),
      callsigns = paste(unique(callsign), collapse = ", "),
      .groups = "drop"
    ) |>
    arrange(desc(n_flights))

  unmatched_our_summary |>
    knitr::kable(
      caption = "Our unmatched flights by route",
      col.names = c("ADEP", "ADES", "Count", "Avg Positions", "Callsigns")
    )
}
```

### Time Distribution: Matched vs Unmatched

```{r}
#| label: plot-matched-vs-unmatched-time
#| fig-width: 12
#| fig-height: 6

# Compare time distribution of matched vs unmatched flights
time_comparison_df <- bind_rows(
  matched_hybrid |>
    mutate(hour = hour(first_time), status = "Matched") |>
    select(adep, ades, hour, status),
  unmatched_our |>
    mutate(hour = hour(first_time), status = "Unmatched") |>
    select(adep, ades, hour, status)
)

ggplot(time_comparison_df, aes(x = hour, fill = status)) +
  geom_histogram(bins = 24, position = "dodge", alpha = 0.7) +
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  scale_fill_manual(values = c("Matched" = "steelblue", "Unmatched" = "coral")) +
  labs(
    title = "Time Distribution: Matched vs Unmatched Flights",
    subtitle = "Flight start time by hour of day",
    x = "Hour of Day (UTC)",
    y = "Number of Flights",
    fill = "Status"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

### Callsign Patterns

```{r}
#| label: callsign-patterns

if (nrow(unmatched_our) > 0) {
  # Extract airline prefix from callsigns
  unmatched_airlines <- unmatched_our |>
    mutate(airline = str_extract(callsign, "^[A-Z]+")) |>
    count(airline, sort = TRUE)

  matched_airlines <- matched_hybrid |>
    mutate(airline = str_extract(callsign, "^[A-Z]+")) |>
    count(airline, sort = TRUE)

  airline_comparison <- full_join(
    matched_airlines, unmatched_airlines,
    by = "airline",
    suffix = c("_matched", "_unmatched")
  ) |>
    replace_na(list(n_matched = 0, n_unmatched = 0)) |>
    mutate(
      total = n_matched + n_unmatched,
      pct_unmatched = round(100 * n_unmatched / total, 1)
    ) |>
    arrange(desc(total))

  airline_comparison |>
    head(10) |>
    knitr::kable(
      caption = "Airline matching success",
      col.names = c("Airline", "Matched", "Unmatched", "Total", "% Unmatched")
    )
}
```

## Visualizations: Example Trajectories

### Matched Flight Examples

```{r}
#| label: plot-matched-examples
#| fig-width: 12
#| fig-height: 8

# Select 3 well-matched flights (callsign match, low time diff)
example_matched <- matched_hybrid |>
  filter(match_method == "callsign") |>
  arrange(time_diff_min) |>
  head(3) |>
  pull(fid)

# Get trajectories
example_matched_traj <- df_trajectories |>
  filter(fid %in% example_matched)

# Plot
ggplot(example_matched_traj) +
  geom_path(aes(x = longitude, y = latitude, group = fid, color = as.factor(fid)),
            linewidth = 1, alpha = 0.8) +
  geom_point(aes(x = longitude_adep, y = latitude_adep),
             color = "green", size = 4, shape = 17) +
  geom_point(aes(x = longitude_ades, y = latitude_ades),
             color = "red", size = 4, shape = 17) +
  facet_wrap(~ paste(adep, "→", ades, "\nFID:", fid), scales = "free") +
  labs(
    title = "Well-Matched Flight Examples",
    subtitle = "Callsign matches with < 1 minute time difference",
    x = "Longitude", y = "Latitude",
    color = "Flight ID"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

### Unmatched Flight Examples

```{r}
#| label: plot-unmatched-examples
#| fig-width: 12
#| fig-height: 8

if (nrow(unmatched_our) >= 3) {
  # Select 3 unmatched flights from different routes
  example_unmatched <- unmatched_our |>
    group_by(adep, ades) |>
    slice_head(n = 1) |>
    ungroup() |>
    head(3) |>
    pull(fid)

  # Get trajectories
  example_unmatched_traj <- df_trajectories |>
    filter(fid %in% example_unmatched)

  # Plot
  ggplot(example_unmatched_traj) +
    geom_path(aes(x = longitude, y = latitude, group = fid, color = as.factor(fid)),
              linewidth = 1, alpha = 0.8) +
    geom_point(aes(x = longitude_adep, y = latitude_adep),
               color = "green", size = 4, shape = 17) +
    geom_point(aes(x = longitude_ades, y = latitude_ades),
               color = "red", size = 4, shape = 17) +
    facet_wrap(~ paste(adep, "→", ades, "\nFID:", fid, "\nCallsign:", callsign), scales = "free") +
    labs(
      title = "Unmatched Flight Examples",
      subtitle = "Flights not found in Brazil HFE reference dataset",
      x = "Longitude", y = "Latitude",
      color = "Flight ID"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
} else {
  cat("Insufficient unmatched flights for visualization")
}
```

### Matching Quality by Route

```{r}
#| label: plot-matching-quality
#| fig-width: 10
#| fig-height: 8

# Calculate matching rate and average time diff by route
route_quality <- matched_hybrid |>
  group_by(adep, ades) |>
  summarise(
    n_matched = n(),
    avg_time_diff = mean(time_diff_min),
    .groups = "drop"
  ) |>
  left_join(
    our_flights |> count(adep, ades, name = "n_total"),
    by = c("adep", "ades")
  ) |>
  mutate(
    match_rate = n_matched / n_total,
    route = paste(adep, "→", ades)
  ) |>
  arrange(desc(n_matched))

ggplot(route_quality, aes(x = reorder(route, match_rate), y = match_rate)) +
  geom_col(aes(fill = avg_time_diff), alpha = 0.8) +
  geom_text(aes(label = sprintf("%d/%d", n_matched, n_total)),
            hjust = -0.2, size = 3) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1.1)) +
  scale_fill_gradient(low = "lightblue", high = "darkblue",
                      name = "Avg Time\nDiff (min)") +
  coord_flip() +
  labs(
    title = "Matching Quality by Route",
    subtitle = "Match rate and average time difference",
    x = "Route",
    y = "Match Rate",
    caption = "Labels show: matched/total flights"
  ) +
  theme_minimal()
```

### Time Difference Distribution by Match Method

```{r}
#| label: plot-time-diff-by-method
#| fig-width: 10
#| fig-height: 6

ggplot(matched_hybrid, aes(x = match_method, y = time_diff_min, fill = match_method)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
  scale_fill_manual(values = c("callsign" = "steelblue", "time" = "coral")) +
  labs(
    title = "Time Difference Distribution by Matching Method",
    subtitle = sprintf("Callsign matches: %.1f min median | Time matches: %.1f min median",
                      median(matched_hybrid$time_diff_min[matched_hybrid$match_method == "callsign"]),
                      median(matched_hybrid$time_diff_min[matched_hybrid$match_method == "time"])),
    x = "Matching Method",
    y = "Time Difference (minutes)",
    fill = "Method"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## FIRST_HIT Timing Analysis

Investigate systematic offset between our FIRST_HIT and Brazil's reference time:

```{r}
#| label: first-hit-timing-analysis

# Calculate signed time difference (positive = our time is later)
matched_timing <- matched_hybrid |>
  mutate(
    signed_time_diff_min = as.numeric(difftime(first_time, bra_time, units = "mins"))
  )

# Summary statistics
timing_stats <- matched_timing |>
  summarise(
    n = n(),
    mean_diff = round(mean(signed_time_diff_min), 2),
    median_diff = round(median(signed_time_diff_min), 2),
    sd_diff = round(sd(signed_time_diff_min), 2),
    our_earlier = sum(signed_time_diff_min < 0),
    bra_earlier = sum(signed_time_diff_min > 0),
    exact_match = sum(abs(signed_time_diff_min) < 0.1)
  )

timing_stats |>
  knitr::kable(
    caption = "FIRST_HIT timing offset analysis",
    col.names = c("N", "Mean Δ", "Median Δ", "SD", "Our Earlier", "BRA Earlier", "Exact Match")
  )
```

**Interpretation**:

- **Negative difference**: Our FIRST_HIT is **before** Brazil's reference time
- **Positive difference**: Our FIRST_HIT is **after** Brazil's reference time

```{r}
#| label: plot-signed-time-diff
#| fig-width: 10
#| fig-height: 6

ggplot(matched_timing, aes(x = signed_time_diff_min)) +
  geom_histogram(bins = 40, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
  geom_vline(xintercept = median(matched_timing$signed_time_diff_min),
             linetype = "solid", color = "darkblue", linewidth = 1) +
  annotate("text", x = median(matched_timing$signed_time_diff_min),
           y = Inf, vjust = 2,
           label = sprintf("Median: %.2f min", median(matched_timing$signed_time_diff_min)),
           color = "darkblue", fontface = "bold") +
  labs(
    title = "FIRST_HIT Timing Offset Distribution",
    subtitle = "Signed difference: Our FIRST_HIT time - Brazil reference time",
    x = "Time Difference (minutes)\n← Our Earlier | Brazil Earlier →",
    y = "Number of Flights",
    caption = "Red line = exact match (0 min) | Blue line = median offset"
  ) +
  theme_minimal()
```

### Timing Offset by Route

```{r}
#| label: plot-timing-by-route
#| fig-width: 10
#| fig-height: 10

route_timing <- matched_timing |>
  mutate(route = paste(adep, "→", ades)) |>
  group_by(route) |>
  mutate(n_route = n()) |>
  ungroup() |>
  filter(n_route >= 3) |>  # Only routes with 3+ flights
  arrange(adep, ades)

ggplot(route_timing, aes(x = reorder(route, signed_time_diff_min, FUN = median),
                         y = signed_time_diff_min)) +
  geom_boxplot(fill = "lightblue", alpha = 0.7, outlier.alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_jitter(width = 0.2, alpha = 0.3, size = 2) +
  coord_flip() +
  labs(
    title = "FIRST_HIT Timing Offset by Route",
    subtitle = "Routes with 3+ matched flights",
    x = "Route",
    y = "Time Difference (minutes)\n← Our Earlier | Brazil Earlier →",
    caption = "Red line = exact match"
  ) +
  theme_minimal()
```

### Timing Offset vs Flight Duration

```{r}
#| label: plot-timing-vs-duration
#| fig-width: 10
#| fig-height: 6

matched_timing_duration <- matched_timing |>
  mutate(
    flight_duration_min = as.numeric(difftime(last_time, first_time, units = "mins"))
  )

ggplot(matched_timing_duration, aes(x = flight_duration_min, y = abs(signed_time_diff_min))) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "loess", color = "red", se = TRUE) +
  labs(
    title = "Timing Offset vs Flight Duration",
    subtitle = "Does longer flight correlate with larger time difference?",
    x = "Flight Duration (minutes)",
    y = "Absolute Time Difference (minutes)"
  ) +
  theme_minimal()
```

### FIRST_HIT vs LAST_HIT Comparison

Check if Brazil's reference time is closer to takeoff or landing:

```{r}
#| label: first-vs-last-hit

# Calculate difference from both FIRST_HIT and LAST_HIT
hit_comparison <- matched_timing |>
  mutate(
    diff_from_first = abs(as.numeric(difftime(first_time, bra_time, units = "mins"))),
    diff_from_last = abs(as.numeric(difftime(last_time, bra_time, units = "mins"))),
    closer_to = if_else(diff_from_first < diff_from_last, "FIRST_HIT", "LAST_HIT")
  )

# Summary
hit_comparison |>
  count(closer_to) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  knitr::kable(
    caption = "Brazil reference time proximity",
    col.names = c("Closer To", "Count", "%")
  )

# Show examples where Brazil time is closer to LAST_HIT
cat("\nFlights where Brazil time is closer to LAST_HIT (landing):\n")
hit_comparison |>
  filter(closer_to == "LAST_HIT") |>
  select(fid, adep, ades, callsign, diff_from_first, diff_from_last) |>
  arrange(desc(diff_from_last)) |>
  head(10) |>
  knitr::kable(
    digits = 2,
    col.names = c("FID", "ADEP", "ADES", "Callsign", "Δ First (min)", "Δ Last (min)")
  )
```

## Summary & Recommendations

**Key Findings**:

1. **Coverage**: Our radar extract captured `r nrow(our_flights)`/`r nrow(bra_flights)` flights (`r round(100*nrow(our_flights)/nrow(bra_flights),1)`%)

2. **Matching Success**:
   - Callsign + Route: `r sum(matched_hybrid$match_method == "callsign")` flights
   - Time-based (15 min): `r sum(matched_hybrid$match_method == "time")` flights
   - **Total**: `r nrow(matched_hybrid)`/`r nrow(our_flights)` (`r round(100*nrow(matched_hybrid)/nrow(our_flights),1)`%)

3. **Time Differences**:
   - Median: `r round(median(matched_hybrid$time_diff_min),1)` minutes
   - Mean: `r round(mean(matched_hybrid$time_diff_min),1)` minutes

4. **Unmatched Flights**: `r nrow(unmatched_our)` (`r round(100*nrow(unmatched_our)/nrow(our_flights),1)`%)
   - Potential reasons: Different callsigns in Brazil dataset, flights outside HFE scope, quality filters

**Why Initial Matching Failed**:

- 60-minute threshold was too strict for many flights
- Did NOT prioritize callsign matching
- Time fields represent different events (FIRST_HIT vs Brazil's reference time)

**Recommendations**:

1. **Use hybrid matching** for production (callsign priority + time fallback)
2. **Investigate specific unmatched callsigns** - may need callsign mapping table
3. **Proceed to HFE validation** with `r nrow(matched_hybrid)` matched flights
4. **Document time reference differences** in methodology

**Next Steps**: Calculate our HFE values and compare against Brazil reference.
